---
title: "Creating episodes"
author: "Göran Broström"
date: "14 maj 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(idsr)
```

## Introduction

This is the **R** version of creating an *episodes* data frame from a 
*chronicle* and an *instruction* frame. The *chronicle* is set up exactly as 
in the Stata version (Quaranta 2015), while the *instruction* frame is an 
extended version of the *VarSetup* file. 

We have a slightly different approach here in that the *chronicle* is supposed 
to contain records of *events*. These events define in turn *changes in levels*
of *variables*, that must be defined.

This approach also forces us to strip off time-fixed covariates, 
such as *sex*, *birth place*, *birth order*, etc, from the *chronicle*. These 
variables are marked *Transition = "Invariant"* in the *instruction*.

The code writing is exemplified with data from DDB.

## The *instruction* data frame

The *instruction* frame must contain the variables (columns) *Type*, 
*Transition*, *Duration*, *variable*, and *mode* (to be explained). Our example 
*instruction* looks like this:

```{r instruction}
knitr::kable(instruction)
```

We split this frame into two frames, *ins_inv*, which contains all time-fixed 
variables, and *ins_events* containing the rest. Note that *Type == variable* in 
*invariant*, so we skip *Type*.

```{r splitinst}
ins_inv <- instruction[instruction$Transition == "Invariant", ]
ins_inv$Type <- ins_inv$Transition <- NULL # Remove ...
ins_events <- instruction[instruction$Transition != "Invariant", ]
```

## The *chronicle* file

The *chronicle* frame must contain the same *Type*s as the *instruction*. It is
an error if this is not fulfilled. If the *chronicle* contains *Type*s not 
present in the *instruction*, the corresponding rows of *chronicle* are removed
with a *warning*.

```{r checktypes}
chronicle <- indi
chronicle$Type[chronicle$Type == "Child_sex"] <- "Sex"
chronicle$Type[chronicle$Type == "Infant_death"] <- "Death"
chronicle$Type[chronicle$Type == "Child_birthdate"] <- "Birthdate"
types <- unique(chronicle$Type)
not_there <- !(instruction$Type %in% types)
if (sum(not_there)){
    cat("The following Type(s) are missing:\n")
    cat(instruction$Type[not_there])
    stop("The instruction must be corrected.")
}
not_there <- !(types %in% instruction$Type)
if (sum(not_there)){
    chronicle <- chronicle[chronicle$Type %in% instruction$Type, ]
    warning("Rows removed in 'chronicle'")
}
```

So, our test data pass the tests and we now split the *chronicle* into *inv*, 
the invariant part, and the rest, *events*.

```{r}
inv <- chronicle[chronicle$Type %in% ins_inv$variable, ]
events <- chronicle[!(chronicle$Type %in% ins_inv$variable), ]
```

### The time invariant frame (*person*)

Some of the *Type*s in the *inv* frame have a time stamp, but no *Value*. For 
those Types, the *Value* is set to the time stamp, in *character* format.

```{r datetypevalue}
datum <- as.Date(paste(inv$Year, inv$Month, inv$Day, sep = "-"))
fill_value <- !is.na(datum)
inv$Value[fill_value] <- as.character(datum[fill_value])
inv$Year <- inv$Month <- inv$Day <- inv$DayFrac <- NULL # Remove ...
```

Now it looks like this (first few records).

```{r lookatinv}
knitr::kable(head(inv))
```

We now transform *inv* to a data frame with new variables given by the content 
in *inv$Type*. The function *spread_* (note the trailing '_') from the package 
*tidyr* does the job.

```{r maketidy}
person <- tidyr::spread_(inv, key_col = "Type", value_col = "Value",
                         convert = TRUE)
```

The *spread_* data frame, *person*, now looks like

```{r}
str(person)
```

It looks OK, but we need to have the right "mode" of the variables. For 
instance, we want the variable "Sex" to be a *factor* with the levels *"Female"*
and *"male"*.

```{r convertinv}
for (nom in ins_inv$variable){
    mode <- ins_inv$mode[ins_inv$variable == nom]
    ##cat("mode = ", mode, "\n")
    if (mode == "date"){
        person[nom] <- as.Date(person[[nom]])
    }else if(mode == "factor"){
        person[nom] <- as.factor(person[[nom]])
    }else{
        person[nom] <- as.vector(person[[nom]], mode = mode)
    }
}
str(person)
```

This looks fine (but note that *Sex* has *three* levels ...).

### The Time varying frame (*events*)

So *events* is summarised as

```{r sumevents}
summary(events)
```

The columns *Year*, *Month*, *Day*, and *Day_Frac* are together forming a *time stamp*, and our first duty is to introduce the **R** *Date* format:

```{r dateformat}
dates <- as.Date(paste(events$Year, events$Month, events$Day, sep = "-"))
events$date <- dates
events$Year <- events$Month <- events$Day <- NULL # remove ...
summary(events)
```

The next step is to add the *DayFrac* to *date* (internally, a *Date* is just a numeric value). Before that we must change *NA* to 0 (zero). Finally *DayFrac* is removed from *events*

```{r adddayfrac}
events$DayFrac[is.na(events$DayFrac)] <- 0
events$date <- events$date + events$DayFrac
events$DayFrac <- NULL # remove ...
knitr::kable(head(events))
```

Next we have to check for duplicates in *events*.

```{r}
dups <- duplicated(events)
(anydup <- sum(dups))
```

So there are no duplicated rows in *events*. Next we create new variables from the *Type* and 
*Value* columns with *tidyr::spread*.

```{r}
episodes <- tidyr::spread_(events, key_col = "Type", value = "Value", convert = TRUE)
knitr::kable(head(episodes))
```


