---
title: "Creating episodes"
author: "Göran Broström"
date: "`r Sys.Date()`"
output: 
   rmarkdown::html_vignette:
      toc: true
      depth: 3
      number_sections: true
vignette: >
  %\VignetteIndexEntry{Creating episodes}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# Introduction

This is the **R** version of creating an *episodes* data frame from a 
*chronicle* and a *personal* frame. The *chronicle* is set up 
similarly to the Stata version (Quaranta 2015).

We have a slightly different approach here in that the *chronicle* is supposed 
to contain records of *events*. These events define in turn *changes in levels*
of *variables*, that must be defined.

This approach also forces us to strip off time-fixed covariates, 
such as *sex*, *birth place*, *birth order*, etc, from the *chronicle*. These 
variables are characterized by the absence of a time-stamp. They are stored in 
the *personal* frame. 

The function must also be told the names of the *start* and *stop* events for 
the aimed-at analysis. We exemplify by studying mortality with birth as start 
event and death as the final event.

The code writing is exemplified with data from Quaranta (2015).

# The *chronicle* file

```{r get_chronichle}
start_event <- "Birth"
end_event <- "Death"
##load("../data/chronicle.rda")
library(idsr)
library(tidyr)
knitr::kable(head(chronicle))
knitr::kable(with(chronicle, table(Variable, Value)))
```

So we have four variables, *alive*, *present*, *civil_status*, and *occupation*.
We "create" those variables with the aid of the function *spread_* in the 
*tidyr* package (note that we do not need the column "Type", so we remove it first):

```{r spread_it}
chron <- chronicle[, c("Id_I", "Variable", "Value", "date", "Type")]
chron <- chron[order(chron$Id_I, chron$date), ]
chron <- chron[!duplicated(chron), ]
chron <- spread_(chron, key_col = "Variable", 
                     value_col = "Value", convert = TRUE)
chron <- dplyr::group_by_(chron, "Id_I")
chron <- tidyr::fill_(chron, names(chron)[-(1:2)])
chron <- dplyr::ungroup(chron)
knitr::kable(chron)
```

We need a "start date", that is, the date of the defining start event (in this 
example it is date of birth).

```{r startdate}
starting <- chron[chron$Type == start_event, c("Id_I", "date")]
indx <- match(chron$Id_I, starting$Id_I)
chron$start_date <- starting$date[indx]
```

Each row in *chronicle* indicates a *start* of an episode, which ends at the 
start of the next spell (for the same individual). The last row for an 
individual is special, in that there is no "next spell", so it needs special
treatment.

We start by calculating duration since *start_date* (time in years). We call
the result *enter*:

```{r}
enter <- as.numeric(chron$date) - as.numeric(chron$start_date)
chron$enter <- round(enter / 365.2425, 3) # 3 decimals is enough
```

We calculate two help variables:

```{r}
indx <- tapply(chron$Id_I, chron$Id_I)
no <- unlist(tapply(chron$Id_I, chron$Id_I, function(x) 1:length(x)))
norec <- tapply(chron$Id_I, chron$Id_I, length)[indx]
```

The last record for each individual is characterized by *no == norec*. So,

```{r addexit}
chron$exit <- c(chron$enter[-1], NA)
chron$exit[no == norec] <- NA
```

Now we need to introduce an indicator of an interval ending with *end_event*. 
Call it *event*. It corresponds to the row that *precedes* the row with 
*Type == end_event*. So,

```{r}
erows <- which(chron$Type == end_event) - 1
chron$event <- FALSE
chron$event[erows] <- TRUE
```

Time to clean up *chron*. 

```{r cleanup}
chron <- chron[!is.na(chron$present) & (chron$present == "yes"), ]
chron <- chron[chron$alive == "yes", ]
remove <- !chron$event & !is.na(chron$exit) & (chron$enter == chron$exit)
chron <- chron[!remove, ]
chron$present <- chron$date <- chron$Type <- chron$alive <- NULL
##chron <- chron[, c("Id_I", "start_date", "enter", "exit", "event", "civil_status", "occupation")]
##chron$civil_status <- factor(chron$civil_status)
##chron$occupation <- factor(chron$occupation)
```


# The *personal* file

The *personal* file looks like this:

```{r personal}
knitr::kable(personal)
```

We *spread* personal and save the result in *per*:

```{r spreadpersonal}
per <- tidyr::spread_(personal, key_col = "Type", value_col = "Value", convert = TRUE)
knitr::kable(per)
```

# Join information

The final step is to put on the fixed-time information from *per* to
*chron* and save the result in *epi* (an episodes file):

```{r joinall}
indx <- match(chron$Id_I, per$Id_I)
vars <- names(per)[-1]
epi <- chron
for (i in vars){
    epi[i] <- per[[i]][indx]
}
knitr::kable(epi)
```

Done!

# Conclusion

This solution put some demand on the input, that is, a great deal has to be done 
in preparing the *chronicle* and the *personal* files. For instance, in this 
example, the *NA*s in the column *civil_status* should maybe changed to
*single* (not married), but how do we know?




