---
title: "From INDIVIDUAL to Chronicle"
author: "Göran Broström"
date: "`r Sys.Date()`"
output: 
   rmarkdown::html_vignette:
      toc: true
      depth: 3
      number_sections: true
vignette: >
  %\VignetteIndexEntry{From  INDIVIDUAL to Chronicle}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction 

This is an example of how to go from an *INDIVIDUAL* file to a *chronicle* file. We use the example data base created by [Luciana Quaranta](http://www.ekh.lu.se/personal/staff/ekh-lqn). It is included in this **R** package.



# Quaranta's Demo Data Base

To make it really simple, with no messy data, we use [Quaranta's demo data base](http://www.ehps-net.eu/article/stata-programs-using-intermediate-data-structure-ids-construct-files-statistical-analysis), which is included in this package as *external data* (an excel file). We convert the first sheet in the file to the **R** data frame *individual*.

```{r loaddemo, message=FALSE}
x <- system.file("extdata", "DemoDatabase.xlsx", mustWork = TRUE, package = "idsr")
library(readxl)
library(dplyr)
individual <- read_excel(x, sheet = 1)
names(individual)
```

It turns out that the variables *Start_day*, *Start_month*, *Start_year*, *End_day*, *End_month*, and *End_year* all contain only missing values (*NA*), so they can safely be removed.  

```{r, tabidd, Source}
with(individual, table(Id_D))
```

Obviously, we can get rid of the *Id_D* column as well:

```{r getrioof}
individual$Start_day <- NULL
individual$Start_month <- NULL
individual$Start_year <- NULL
individual$End_day <- NULL
individual$End_month <- NULL
individual$End_year <- NULL
individual$Id_D <- NULL
##individual <- individual %>%
  ##  select(Id_I, Type, Value, Value_Id_C, Day, Month, Year, Date_type, Source) %>%
    ##arrange(Id_I, Year, Month, Day)
knitr::kable(head(individual, 8))
```


This is the full information available for individual No. `r as.integer(individual$Id_I[1])`.

## Properties of a chronicle frame

The chronicle frame should only contain variables corresponding to *Date_type = Event*. However, this is too restrictive, because for some reason events of certain types are *not* reported as *Event*. From a strict database-management perspective that is reasonable, but not in the afterworld of data (survival) analysis. Therefore I let all rows with a (correct) date be part of the chronicle frame.

The other variables are time-constant and we keep them separate from the rest: They are very easy to add on at the end. There are two exceptions: The *Declared* Types *Start_observation* and *End_observation* need to be in the chronicle frame

```{r splitup}
##to_chron <- (individual$Date_type == "Event") | 
  ##          (individual$Type %in% c("Start_observation", "End_observation"))
to_chron <- !is.na(individual$Year)
personal <- individual[!to_chron, ]
chronicle <- individual[to_chron, ]
chronicle$Date_Type <- NULL # Not needed
```

So, the *chronicle* frame looks like this at the start:


```{r chronstart}
knitr::kable(chronicle)
```

We obvously don't need the variables *Value_Id_C* and *Source* in *chronicle*, so we remove them. The variable *Date_type* can also be removed, since it is constant (*Event*) in this frame.
We also convert the *Year-Month-Year* triple into a *Date* variable, after wich we can remove the triple. After that the frame is sorted by *Id_I*,*date*, and in case of tied dates, *Type == Start_observation* is first and *Type == End_observation* is last.

```{r}
##chronicle$Value_Id_C <- NULL
##chronicle$Source <- NULL
##chronicle$Date_type <- NULL
chronicle$date <- as.Date(paste(chronicle$Year, 
                                chronicle$Month, 
                                chronicle$Day, sep = "-"))
##chronicle$Year <- NULL
##chronicle$Month <- NULL
##chronicle$Day <- NULL
chronicle <- chronicle[, c("Id_I", "Type", "Value", "date")]
chronicle <- chronicle[order(chronicle$Id_I,
                             chronicle$date,
                             chronicle$Type != "Start_observation",
                             chronicle$Type == "End_observation"), ]
knitr::kable(chronicle)
```

Very neat!

Now, for each distinct *Type* (event) we need to define a *variable*: The events *Birth* and *Death* corresponds to the logical variable *alive* (becomes *TRUE* at a birth and *FALSE* at a death). The event *Marriage* corresponds to the variable *civil_status* (becomes *married* at a marriage, and is *unmarried* at start).

So,

```{r insertvarr}
chronicle$Variable <- NA
take <- chronicle$Type %in% c("Birth", "Birth_date")
chronicle$Variable[take] <- "alive"
chronicle$Value[take] <- "yes"
##
take <- chronicle$Type %in% c("Death", "Death_date")
chronicle$Variable[take] <- "alive"
chronicle$Value[take] <- "no"
##

take <- chronicle$Type == "Start_observation"
chronicle$Variable[take] <- "present"
chronicle$Value[take] <- "yes"
take <- chronicle$Type == "End_observation"
chronicle$Variable[take] <- "present"
chronicle$Value[take] <- "no"
##
chronicle$Variable[chronicle$Type == "Occupation"] <- "occupation"
##
take <- chronicle$Type == "Marriage"
chronicle$Variable[take] <- "civil_status"
chronicle$Value[take] <- "married"
## Reorder:
chronicle <- chronicle[, c("Id_I", "Type", "Variable", "Value", "date")]
knitr::kable(chronicle)

```


We also note that not all individuals have a *Birth* event (observed). However, all have a *Birth_date*, in the *personal* frame.  Let us look at it.

## The *personal* frame

The *personal* frame is made tidy in a similar manner as the chronicle frame.

```{r tidypersonal}
personal <- personal[, c("Id_I", "Type", "Value", "Value_Id_C")]
knitr::kable(personal)
```



# Conclusion

So, the input to an *"EpisodesFileCreator"* should be the frames *chronicle* 
and *personal*. In addition, we need a description frame linking *events* to 
*variables* in the *chronicle* frame. See the *episodes* vignette for the continuation.

# References

Alter, George & Kees Mandemakers, 'The Intermediate Data Structure (IDS) for Longitudinal Historical Microdata, version 4', *Historical Life Course Studies* 1 (2014), 1--26.
[http](http://hdl.handle.net/10622/23526343-2014-0001?locatt=view:master)

Quaranta, Luciana, 'Using the Intermediate Data Structure (IDS) to Construct Files for Statistical Analysis', *Historical Life Course Studies* 2 (2015), 86--107.
[http](http://hdl.handle.net/10622/23526343-2015-0007?locatt=view:master)

Quaranta, Luciana, 'Stata Programs for Using the Intermediate Data Structure (IDS) to Construct files for Statistical Analysis', *Historical Life Course Studies* 3 (2015), 1--19. [http](http://hdl.handle.net/10622/23526343-2016-0001?locatt=view:master())
