---
title: "From INDIVIDUAL to Chronicle"
author: "Göran Broström"
date: "`r Sys.Date()`"
output: 
   rmarkdown::html_vignette:
      toc: true
      depth: 3
      number_sections: true
vignette: >
  %\VignetteIndexEntry{From  INDIVIDUAL to Chronicle}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

# Introduction 

This is an example of how to go from an *INDIVIDUAL* file to a *chronicle* file. We use the example data base created by [Luciana Quaranta](http://www.ekh.lu.se/personal/staff/ekh-lqn). It is included in this **R** package.



# Quaranta's Demo Data Base

To make it really simple, with no messy data, we use [Quaranta's demo data base](http://www.ehps-net.eu/article/stata-programs-using-intermediate-data-structure-ids-construct-files-statistical-analysis), which is included in this package as *external data* (an excel file). We convert the first sheet in the file to the **R** data frame *individual*.

```{r loaddemo, message=FALSE}
x <- system.file("extdata", "DemoDatabase.xlsx", mustWork = TRUE, package = "idsr")
library(readxl)
library(dplyr)
individual <- read_excel(x, sheet = 1)
names(individual)
```

It turns out that the variables *Start_day*, *Start_month*, *Start_year*, *End_day*, *End_month*, and *End_year* all contain only missing values (*NA*), so they can safely be removed.  

```{r, tabidd, Source}
with(individual, table(Id_D))
```

Obviously, we can get rid of the *Id_D* column as well:

```{r getrioof}
individual$Start_day <- NULL
individual$Start_month <- NULL
individual$Start_year <- NULL
individual$End_day <- NULL
individual$End_month <- NULL
individual$End_year <- NULL
individual$Id_D <- NULL
##individual <- individual %>%
  ##  select(Id_I, Type, Value, Value_Id_C, Day, Month, Year, Date_type, Source) %>%
    ##arrange(Id_I, Year, Month, Day)
knitr::kable(head(individual, 8))
```


This is the full information available for individual No. `r as.integer(individual$Id_I[1])`.

## Properties of a chronicle frame

The chronicle frame should only contain variables corresponding to *Date_type = Event*. However, this is too restrictive, because for some reason events of certain types are *not* reported as *Event*. From a strict database-management perspective that is reasonable, but not in the afterworld of data (survival) analysis. Therefore I let all rows with a (correct) date be part of the chronicle frame.

The other variables are time-constant and we keep them separate from the rest: They are very easy to add on at the end. There are two exceptions: The *Declared* Types *Start_observation* and *End_observation* need to be in the chronicle frame

```{r splitup}
##to_chron <- (individual$Date_type == "Event") | 
  ##          (individual$Type %in% c("Start_observation", "End_observation"))
to_chron <- !is.na(individual$Year)
personal <- individual[!to_chron, ]
chronicle <- individual[to_chron, ]
chronicle$Date_Type <- NULL # Not needed
```

So, the *chronicle* frame looks like this at the start:


```{r chronstart}
knitr::kable(chronicle)
```

We obvously don't need the variables *Value_Id_C* and *Source* in *chronicle*,
so we remove them. The variable *Date_type* can also be removed, since it is 
constant (*Event*) in this frame.
We also convert the *Year-Month-Year* triple into a *Date* variable, after wich 
we can remove the triple. After that the frame is sorted by *Id_I*,*date*, and 
in case of tied dates, *Type == Start_observation* is first and 
*Type == End_observation* is last.

```{r}
##chronicle$Value_Id_C <- NULL
##chronicle$Source <- NULL
##chronicle$Date_type <- NULL
chronicle$date <- as.Date(paste(chronicle$Year, 
                                chronicle$Month, 
                                chronicle$Day, sep = "-"))
##chronicle$Year <- NULL
##chronicle$Month <- NULL
##chronicle$Day <- NULL
chronicle <- chronicle[, c("Id_I", "Type", "Value", "date")]
chronicle <- chronicle[order(chronicle$Id_I,
                             chronicle$date,
                             chronicle$Type != "Start_observation",
                             chronicle$Type == "End_observation"), ]
knitr::kable(chronicle)
```

Very neat! However, for a specific study we need a well-defined *start event* 
and a likewise well-defined *end event*. We exemplify by thinking of 
*mortality*: Following individuals from birth to death. So our start event is 
*Birth*, and it must be defined (including a date) for all individuals. The end
event is *Death*, but its date is not necessarily known to us 
("right censoring"). Note the difference: We may not observe a birth in our 
data, but its date must be known.

For our purpose, *Birth* and *Birth_date* carry the same information in the 
presence of the event *Start_observation*, which is mandatory. So we replace all 
*Type = Birth_date* by *Type = Birth* and then remove duplicates:

```{r dupbirths}
chronicle$Type[chronicle$Type == "Birth_date"] <- "Birth"
chronicle <- chronicle[!duplicated(chronicle), ]
```

Next we must define two new logical variables, *start_event* (TRUE exactly once for all individuals) and *end_event* (TRUE at most once for all). However, it s more practical to defer that exercise to the episodes file creation.


Now, for each distinct *Type* (event) we need to define a *variable*: The events *Birth* and *Death* corresponds to the logical variable *alive* (becomes *TRUE* at a birth and *FALSE* at a death). The event *Marriage* corresponds to the variable *civil_status* (becomes *married* at a marriage, and is *unmarried* at start).

So,

```{r insertvarr}
chronicle$Variable <- NA
take <- chronicle$Type %in% c("Birth", "Birth_date")
chronicle$Variable[take] <- "alive"
chronicle$Value[take] <- "yes"
##
take <- chronicle$Type %in% c("Death", "Death_date")
chronicle$Variable[take] <- "alive"
chronicle$Value[take] <- "no"
##

take <- chronicle$Type == "Start_observation"
chronicle$Variable[take] <- "present"
chronicle$Value[take] <- "yes"
take <- chronicle$Type == "End_observation"
chronicle$Variable[take] <- "present"
chronicle$Value[take] <- "no"
##
chronicle$Variable[chronicle$Type == "Occupation"] <- "occupation"
##
take <- chronicle$Type == "Marriage"
chronicle$Variable[take] <- "civil_status"
chronicle$Value[take] <- "married"
## Reorder:
chronicle <- chronicle[, c("Id_I", "Variable", "Value", "date", "Type")]
knitr::kable(chronicle)
```

There is one problem remaining: Duplicates. Let us look at individual 
*Id_I = 1548468*:

```{r iddup}
knitr::kable(chronicle[chronicle$Id_I == 1548468, ])
```

Two problems:

1. Two birth notifications on the same date (thank heaven). Easy to fix: Remove
   one of them.

2. Death on the same day as birth. Must keep both (of course), but we must 
   decide which comes first. That is also easy, birth before death. Could be 
   solved by 
   introcucing a *DayFrac*, meaning that we add a small number (0.02, say)to 
   the death date, but we avoid that, at least for the moment. 
   
   


```{r dayfrac}
chronicle <- chronicle[, c("Id_I", "Variable", "Value", "date", "Type")]
chronicle <- chronicle[!duplicated(chronicle), ]
##rows <- duplicated(chronicle[, c("Id_I", "Variable", "date")])   # NOTE:
##chronicle$date[rows] <- chronicle$date[rows] + 0.02     # Outcommented!!
```

## The *personal* frame

The *personal* frame is made tidy in a similar manner as the chronicle frame.

```{r tidypersonal}
personal <- personal[, c("Id_I", "Type", "Value", "Value_Id_C")]
knitr::kable(personal)
```

The column *Value_Id_C* and the rows with *Type == Birth_location* are without
useful information, so they are removed:

```{r rempers}
personal$Value_Id_C <- NULL
personal <- personal[personal$Type != "Birth_location", ]
knitr::kable(personal)
```


# Conclusion

So, the input to an *"EpisodesFileCreator"* should be the frames *chronicle* 
and *personal*. In addition, we need a description frame linking *events* to 
*variables* in the *chronicle* frame. See the *episodes* vignette for the continuation.

# References

Alter, George & Kees Mandemakers, 'The Intermediate Data Structure (IDS) for Longitudinal Historical Microdata, version 4', *Historical Life Course Studies* 1 (2014), 1--26.
[http](http://hdl.handle.net/10622/23526343-2014-0001?locatt=view:master)

Quaranta, Luciana, 'Using the Intermediate Data Structure (IDS) to Construct Files for Statistical Analysis', *Historical Life Course Studies* 2 (2015), 86--107.
[http](http://hdl.handle.net/10622/23526343-2015-0007?locatt=view:master)

Quaranta, Luciana, 'Stata Programs for Using the Intermediate Data Structure (IDS) to Construct files for Statistical Analysis', *Historical Life Course Studies* 3 (2015), 1--19. [http](http://hdl.handle.net/10622/23526343-2016-0001?locatt=view:master())
